\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Variable Models for Networks

### 2018-04-23 manuscript_rearranged.tex
revising description

 \item Simulate $\mathbf{x} = \frac{\rho}{\sqrt{1-\rho^2}}\mathbf{d}+\mathbf{w},$ where $\mathbf{w} \sim N(0,1)$. This assures that the correlation between $\mathbf{x}$ and $\mathbf{d}$ is $\rho$.\footnote{This could be generalized to more than one covariate using the conditional normal distribution derived by \cite[pp. 116--117]{eaton1983}.}

 \item Finally we standardize $\mathbf{x} := \mathbf{x}/ \text{sd}(\mathbf{x})$ to have unit variance to ensure that the variance of $\mathbf{x}$ relative to $\mathbf{d}$ is unrelated to $\rho$. This is to assure that $\rho$ does not affect the efficiency with which the effect of $\mathbf{x}$ can be estimated through some property unrelated to the relationship between $\mathbf{x}$ and $\mathbf{d}$.

@@

We consider the LSM with several different priors on the coefficient
for the observed covariate and the latent space. We set the prior
variance of $\beta$ to be either $1$ or $10$ and use a diffuse normal
prior on the latent space.

The LSM is estimated using the canonical implementation in the
\texttt{latentnet} package in \texttt{R} \citep{latentnet}. For each
iteration of the simulation, we run 10,000 burn in iterations,
followed by 1,000,000 iterations of the sampler. Every 100th post burn
in iteration is saved. Convergence in the log probability of the model
is assessed using the Geweke diagnostic in the \texttt{coda} package
\citep{coda, geweke1991evaluating}. If the convergence criterion is
satisfied, the simulation continues to the next set of arguments,
otherwise the number of iterations is doubled. If the convergence
criterion is still not satisfied, then the aforementioned step in the
simulation is flagged for review. However, this was not necessary in
any cases. At each point in the simulation's parameter space, we
execute 1,000 Monte Carlo iterations. \footnote{One of the primary
  difficulties in executing the above simulation design is the
  computational cost of estimating the LSM. We utilize the
  \texttt{BatchExperiments} \R package to construct and execute our
  computational experiments on a Torque cluster
  \cite{bischl2015batchjobs}.}

We compute the MLE estimated by iteratively reweighted least squares
via the \texttt{glm} function in \texttt{R} \citep{rcore}. For the LSM
we use the posterior mode as our point estimate. In the cases where
edges are Bernoulli and there is omitted network structure, we scale
the estimates using the reciprocal of the bias of $\beta$ when
$\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 +
    \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the
variance of a standard logistic distribution. We do this to adjust for
bias in the coefficient estimate that arises due to the lack of a
scale coefficient in logistic regression (See the derivation of the
bias under an independent but omitted covariate in
\cite{mood2010logistic}).

@@

We also repeat this simulation study using \texttt{amen}
\citep{hoff2015dyadic, minhas2016inferential}. We omit the Poisson
family, and, due to the computational cost of the simulation, we
consider only networks with 25 nodes or 100 nodes. Rather than
constructing an omitted variable as described above, we draw from a
standard normal distribution $\mathbf{u}$, and, because we are
considering symmetric networks, take the outer product of $\mathbf{u}$
with itself to construct the latent factor. We take 100,000 MCMC
samples after 500 burn-in iterations, and we assess convergence using
the diagnostic of \cite{raftery1992practical}.

@@

To evaluate estimation error we compute the bias. Figures
\ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the
results. In Figure \ref{fig:estimation_ls} we can see that with a
uniform distribution over the correlation between the distances and
the observed covariate, all methods (except the true model wherein the
distances are included via an observed covariate) show substantial
bias. In most cases the LSM performs better than the GLM, sometimes by
as much as $50\%$. However, in absolute terms, the amount of bias is
large. Our results demonstrate that using the LSM will not permit the
discovery of and adjustment for the true latent positions when the
observed covariates and latent distances are confounded.  For both the
GLM and the LSM, bias decays rapidly as the degree of correlation
between the distances and the measured covariate decreases. In
\ref{fig:estimation_nls} we can see that when there is no omitted
network structure, the LSM exhibits bias greater than that of the GLM.

@@

\newpage
\section*{Online Appendix}
Below we replicate the simulation study results figures using the results from our simulation with the recently developed \texttt{amen} package. Note that these simulations exclude (1) the Poisson family, which is not available in the \texttt{amen} package, and (2) the 50-node networks, which we left out in the interest of lessening the computational burden of the simulation.

>>>>

 \item Simulate the observed covariate $\mathbf{x} = \frac{\rho}{\sqrt{1-\rho^2}}\mathbf{d}+\mathbf{w},$ where $\mathbf{w} \sim N(0,1)$. This assures that the correlation between $\mathbf{x}$ and $\mathbf{d}$ is $\rho$.\footnote{This could be generalized to more than one covariate using the conditional normal distribution derived by \cite[pp. 116--117]{eaton1983}.}

 \item Finally, we standardize $\mathbf{x} := \mathbf{x}/ \text{sd}(\mathbf{x})$ to have unit variance to ensure that the variance of $\mathbf{x}$ relative to $\mathbf{d}$ is unrelated to $\rho$. This is to assure that $\rho$ does not affect the efficiency with which the effect of $\mathbf{x}$ can be estimated through some property unrelated to the relationship between $\mathbf{x}$ and $\mathbf{d}$.

@@

We consider the LSM with several different priors on the coefficient
for the observed covariate ($\mathbf{x}$) and the latent space. We set the prior
variance of $\beta$ to be either $1$ or $10$ and use a diffuse normal
prior on the latent space.

The LSM is estimated using the canonical implementation in the
\texttt{latentnet} package in \texttt{R} \citep{latentnet}. For each
iteration of the simulation, we run 10,000 burn in iterations of the Metropolis-Hastings sampler in \texttt{latentnet},
followed by 1,000,000 iterations of the sampler. Every 100th post burn
in iteration is saved. Convergence in the log probability of the model
is assessed using the Geweke diagnostic in the \texttt{coda} package
\citep{coda, geweke1991evaluating}. If the convergence criterion is
satisfied, the simulation continues to the next set of arguments,
otherwise the number of iterations is doubled. If the convergence
criterion were still not satisfied, then the aforementioned step would be flagged for review. However, this was not necessary in
any cases---all converged within 3,000,000 iterations. At each point in the simulation's parameter space, we
execute 1,000 Monte Carlo iterations. \footnote{One of the primary
  difficulties in executing the above simulation design is the
  computational cost of estimating the LSM. We utilize the
  \texttt{BatchExperiments} \R package to construct and execute our
  computational experiments on a Torque cluster
  \cite{bischl2015batchjobs}.}

We compute the MLE estimated by iteratively reweighted least squares
via the \texttt{glm} function in \texttt{R} \citep{rcore}. For the LSM
we use the posterior mode as our point estimate. In the cases where
edges are Bernoulli and there is omitted network structure, we scale
the estimates using the reciprocal of the bias of $\beta$ when
$\mathbf{x}$ and $\mathbf{d}$ are independent: $\sqrt{\frac{3.28 +
    \beta^2 \text{Var}(\mathbf{d})}{3.29}}$, where $3.29$ is the
variance of a standard logistic distribution. We do this to adjust for
bias in the coefficient estimate that arises due to the lack of a
scale coefficient in logistic regression (See the derivation of the
bias under an omitted orthogonal covariate in
\cite{mood2010logistic}).

@@

@@

To evaluate estimation error we compute the bias. Figures
\ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the
results. In Figure \ref{fig:estimation_ls} we can see that with a
uniform distribution over the correlation between the distances and
the observed covariate, all methods (except the true model wherein the
distances are included via an observed covariate) show substantial
bias. In most cases the LSM performs better than the GLM, sometimes by
as much as $50\%$. However, in absolute terms, the amount of bias is
large. Our results demonstrate that using the LSM will not permit the
discovery of and adjustment for the true latent positions when the
observed covariates and latent distances are confounded.  For both the
GLM and the LSM, bias decays rapidly as the degree of correlation
between the distances and the measured covariate decreases. In
\ref{fig:estimation_nls} we can see that when there is no omitted
network structure, the LSM exhibits bias greater than that of the GLM.\footnote{ We also repeat this simulation study using \texttt{amen}
\citep{hoff2015dyadic, minhas2016inferential,amen},  which is a software implementation of a much more recently developed algorithm for latent variable modeling of networks. The results of the simulations using \texttt{amen} , which are in agreement with the results with \texttt{latentnet}, are in the appendix.}

@@

\newpage
\section*{Online Appendix}
Below we replicate the simulation study results figures using the results from our simulation with the recently developed \texttt{amen} package.  The latent variable is constructed differently in \texttt{amen} than in the LSM.  Rather than
taking the Euclidean distance between two nodes' latent variables, in the simulation with \texttt{amen} we use the outer product between two nodes' latent variables (just the scalar product in our case, since the latent variable has one dimension). Since inference with \texttt{amen} tends to converge much faster than with the LSM, we take 100,000 MCMC
samples after 500 burn-in iterations, and we assess convergence using
the diagnostic of \cite{raftery1992practical}. Note that these simulations exclude (1) the Poisson family, which is not available in the \texttt{amen} package, and (2) the 50-node networks, which we left out in the interest of lessening the computational burden of the simulation.
